---
title: "Summary Statistics for Loan Portfolio (Cartera de Credito) Data"
author: "Jesús López-Pérez"
date: "18/4/2022"
output: pdf_document
urlcolor: blue
---

```{r setup, echo=FALSE, message=FALSE, cache = FALSE}
knitr::opts_chunk$set(dev='png')

options(digits = 3)

# set up knitr global chunk options
knitr::opts_chunk$set(fig.width = 8)

RStudioView <- View
View <- function(x) {
  name  <- deparse(substitute(x))
  if ("data.frame" %in% class(x)) { 
    RStudioView(x[1:1000,], name)
  } else { 
    RStudioView(x) 
  }
}


```

In this document we present an exploratory data analysis for the data *Información de Cartera de Crédito, Cartera comercial* for Mexican banks downloaded from [CNBV portfolio](https://portafolioinfo.cnbv.gob.mx/Paginas/Reporte.aspx?s=40&t=2&st=3&ti=0&sti=0&n=0&tp=0).

```{r library-imports, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr)
library(ggplot2)
library(zoo)
library(fpp3)
library(tibble)
library(Hmisc)
library(tidyr)
library(patchwork)
library(plm)

```

```{r preparation-and-data-import, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE}
library(tidyverse)
library(stringr)
library(ggplot2)
library(zoo)
library(fpp3)
library(tibble)
library(Hmisc)
library(tidyr)
# rm(list = ls())
Sys.setlocale("LC_TIME", "C")
options(scipen=1, digits=3)

# to prevent viewing large datasets
# https://stackoverflow.com/questions/48792084/view-function-showing-all-columns-makes-r-really-slow


path <- "D:/Personal/MTE/Tesis/Propuestas_Temas/credit-markets/Data"

which_db <- "ALL"
 
# catalogs
read.csv(paste(path, "/cat_ents.csv", sep = ""), encoding = "UTF-8",
         col.names = c("state_code","state_long", "state_short")) %>% 
  mutate(state_code = str_pad(state_code, pad = "0", width = 2, side = "left")) ->  cat_ents

cat_ents <- cat_ents %>% 
  mutate(state_code = ifelse(state_code == 999,99,state_code))

read.csv(paste(path, "/cat_eeffs.csv", sep = ""), encoding = "UTF-8",
         col.names = c("institution_code", "institution")) %>% 
  mutate(institution_code = as.character(str_pad(institution_code,
                                                 pad = "0", width = 6, side = "left"))) -> cat_eeffs

read.csv(paste(path, "/cat_sectores.csv", sep = ""), encoding = "UTF-8",
         col.names = c("sector_code", "Econ_Act", "scian_code", "scian_code2",
                       "description")) %>% 
  mutate(sector_code = as.character(str_pad(sector_code,
                                            pad = "0", width = 2, side = "left"))) -> cat_sects

read.csv(paste(path, "/cat_destino.csv", sep = ""), encoding = "UTF-8",
         col.names = c("funds_use_cnbv_code", "use_description","funds_use_code")) -> cat_funds_use

read.csv(paste(path, "/cat_reporte_cnbv.csv", sep = ""), encoding = "UTF-8",
         col.names = c("cnbv_report_code","report_description",
                       "large_description")) -> cat_cnbv_report

read.csv(paste(path, "/cat_firm_size.csv", sep = ""), encoding = "UTF-8",
         col.names = c("firm_size_code", "firm_description")) -> cat_firm_size

read.csv(paste(path, "/cat_currency.csv", sep = ""), encoding = "UTF-8",
         col.names = c("currency_code", "currency")) -> cat_currency

# read data
# data_r <- read.csv(paste(path, "/proc/firms_G7_2T_16_4T_21.csv", sep = ""),
#                    fileEncoding = "UTF-8")

banks <- c("Citibanamex","BBVA México","Santander","HSBC","Inbursa",
           "Scotiabank","Banorte")

data_r_all <- read.csv(paste(path, "/proc/firms_",which_db,"_2T_16_4T_21.csv", sep = ""),
                   fileEncoding = "UTF-8")
data_r_all %>% 
  mutate(institution_code = str_pad(institution_code,width = 6, pad = "0", side = "left"),
         state_code = str_pad(state_code,width = 2, pad = "0", side = "left")) -> data_r_all

months <- unique(data_r_all$period)
all_banks  <- unique(data_r_all$institution)

```

## Summary of data

The raw data is monthly from **`r months[1]`** to **`r months[length(months)]`** (`r length(months)`) periods. For every month, we have loans data at firm level (one firm can have one or more register (RFC), and each register can have one or more credits).

## Description of variables:

For each period `t`

-   Firm information
    -   `institution_code`: institution `b` that granted loan `r length(all_banks)` banks.
    -   `cnbv_report_code`: `r length(unique(data_r_all$cnbv_report_code))` report types
    -   `scian_code`: `r length(unique(data_r_all$scian_code))` sectors.
    -   `GA_code`: `r length(unique(data_r_all$GA_code))` Grand Economic Activities (primary, secondary and tertiary sector).
    -   `state_code`: `r length(unique(data_r_all$state_code))-1` states + Foreign.
    -   `funds_use_code`: `r length(unique(data_r_all$funds_use_code))` funds use.
    -   `currency_code`: `r length(unique(data_r_all$currency_code))`, Moneda Nacional, Moneda Extranjera, UDIS .
    -   `firm_size_code`, 3 sizes: S - Small, L - Large, T - Trustee.
-   Loan information
    -   `number_of_loans`: Number of loans granted to firm `i`
    -   `number_of_debtors_by_rfc`: Number of debtors within a registered firm `i` (RFC).
    -   `loan_disposition_month`: Loan amount required in month `t`.
    -   `total_portfolio`: Total amount of loans (cummulative).
    -   `current_portfolio`: Total amount of loans with no arrears (cummulative).
    -   `past_due_portfolio`: Total amount of loans with unpaid payments (cummulative).
    -   `interest_rate_disposition`: Calculated interest rate of the amount required in month `t` (int_rate_x\_loan_amount / loan_disposition_month).
    -   `interest_rate_all`: Calculated interest rate for all amount (int_rate_x\_total_resp / total_portfolio).
    -   `term_disposition`: Calculated loan term of the amount required in month `t` (term_x\_loan_amount / loan_disposition_month).
    -   `term_all`: Calculated loan term for all amount (term_x\_total_resp / total_portfolio).
    -   `average_loan`: Calculated average loan amount (total_portfolio / number_of_loans)
    -   `past_due_pct`: Calculated amount of loans with unpaid payments (past_due_portfolio / total_portfolio).

The database contains `r nrow(data_r_all`` rows,

The data does not includes a firm id number, thus I generated a `firm_id` by concatenating:

-   institution_code,# 6 digits
-   cnbv_report_code, \# 3 digits
-   scian_code2,# 2 digits
-   state_code,# 2 digits
-   funds_use_code,# 3 digits
-   currency_code, \# 2 letters
-   firm_size_code, \# 1 letter

example: `r data_r_all$firm_id[1]` stand for firm with bank `040002` from report `R443` Empresas menores 14mdu, sector `11` Farming, state `01` Aguascalientes, `100` funds use Cartas de crédito, `MN` in local currency and `L` large sized firm.

The unique `firm_id`'s are `r  data_r_all %>%  summarise(N_firms_aprox = length(unique(firm_id)))`, however there is no guarantee that these id's belong to unique firms, there are firms with the same code. Thus, we cannot use the data at a firm level.

```{r loan-id-analysis,warning=FALSE, message=FALSE, comment = "", echo =FALSE}
 
data_r_all %>% group_by(firm_id) %>% 
  summarise(N = n()) %>% 
  mutate(N_rep = case_when((N > 0 & N <= 10) ~ "1 to 10 obs",
                          (N >10 & N <= 20) ~ "11 to 20 obs",
                          (N >20 & N <= 30) ~ "21 to 30 obs",
                          (N >30 & N <= 40) ~ "31 to 40 obs",
                          (N >40 & N <= 50) ~ "41 to 50 obs",
                          (N >50 & N <= 60) ~ "51 to 60 obs",
                          (N >60 & N <= 68) ~ "61 to 68 obs",
                           N > 68 ~ "More than 68 obs")) %>% 
  mutate(N_rep = fct_relevel(N_rep,"1 to 10 obs",
                             "11 to 20 obs","21 to 30 obs",
                             "31 to 40 obs","41 to 50 obs",
                             "51 to 60 obs","61 to 68 obs",
                             "More than 68 obs") )%>% 
  group_by(N_rep) %>% 
  summarise(count_id = sum(N)) -> table_id_count #%>% pull(count_id) %>% sum()

knitr::kable(table_id_count,digits = 2, format.args = list(big.mark = ",",   scientific = FALSE)) 

table_id_count %>% 
  ggplot(aes(x = N_rep, y = count_id)) + 
  geom_bar(stat = "identity") ->g1


```

## Data cleaning

We remove loans with zero interest rate that represent

```{r data-cleaning, comment="", message=FALSE, echo = FALSE, warning=FALSE}

nrow(data_r_all)

# we remove:
data_r_all %>% filter(interest_rate_disposition <= 0) %>% nrow # 17710
data_r_all %>% filter(interest_rate_all <= 0) %>% nrow # 14949
data_r_all %>% filter(term_disposition < 0) %>% nrow # 36
data_r_all %>% filter(term_all < 0) %>% nrow # 7663
data_r_all %>% filter(term_all > 360) %>% nrow # 19035

# temp <- ifelse(data_r_all$interest_rate_disposition <0,"Yes","No")
# length(which(temp == "Yes"))

# 
clean <- function(x, include = FALSE){
    x[is.na(x)] <- include
    x
}

# # Original output
# with(data_r_all, data_r_all[interest_rate_disposition <0 ,])
# # Clean it up and remove NAs
# with(data_r_all, data_r_all[clean(interest_rate_disposition <0),])
# # Clean it up but include NAs


# data_r_all$ID = cumsum(!is.na(data_r_all$interest_rate_disposition))
# data_r_all$ID <- ifelse(is.na(data_r_all$interest_rate_disposition),
#                       data_r_all$interest_rate_disposition,data_r_all$ID)
# 

# cleaning interest rates
N <- nrow(data_r_all)
n_orig <- nrow(data_r_all)
data_r_all$log_int_rate_d <- log(data_r_all$interest_rate_disposition)
data_r_all$log_int_rate_a <- log(data_r_all$interest_rate_all)
data_r_all <- with(data_r_all, data_r_all[clean(log_int_rate_d >=0, include = TRUE),])
data_r_all <- with(data_r_all, data_r_all[clean(log_int_rate_a >=0, include = TRUE),])
data_r_all <- with(data_r_all, data_r_all[clean( !is.infinite(interest_rate_all), include = TRUE),])
data_r_all <- with(data_r_all, data_r_all[clean( !is.infinite(interest_rate_disposition), include = TRUE),])
n_clean1 <- nrow(data_r_all)
n_orig - n_clean1; 100* (n_orig - n_clean1) / n_orig 
n_removed_int <- n_orig - n_clean1

# cleaning term
data_r_all <- with(data_r_all, data_r_all[clean(term_all >=0, include = TRUE),])
data_r_all <- with(data_r_all, data_r_all[clean(term_disposition >=0, include = TRUE),])
data_r_all <- with(data_r_all, data_r_all[clean( !is.infinite(term_all), include = TRUE),])
data_r_all <- with(data_r_all, data_r_all[clean( !(term_all > 360), include = TRUE),]) 
data_r_all <- with(data_r_all, data_r_all[clean( !(term_disposition > 360), include = TRUE),])
n_clean2 <- nrow(data_r_all)
n_orig  - (n_clean2 ); 100* (n_orig - (n_clean2)) / n_orig 
n_clean2


nrow(data_r_all)

write.csv(data_r_all, paste(path, "/proc/firms_ALL_2T_16_4T_21-clean.csv", 
                            sep = ""),row.names = FALSE,
          fileEncoding = "UTF-8")

```

## Summary statistics for numerical variables

```{r summary, echo=FALSE, comment="", message=FALSE, warning=FALSE}
library(tidyr)
library(tidyverse)


data_r_all %>% select(number_of_loans, 
                      number_of_debtors_by_rfc, 
                      loan_disposition_month,
                      interest_rate_disposition,
                      term_disposition) %>% 
  apply(2,fivenum) -> summary_1

summary_1 <- t((summary_1))
colnames(summary_1) <- c("min", "p25","median","p75","max")
 knitr::kable(summary_1,digits = 2, format.args = list(big.mark = ",",   scientific = FALSE))
                      
data_r_all %>% select(total_portfolio, 
                      current_portfolio,
                      average_loan,
                      past_due_portfolio,
                      past_due_pct,
                      interest_rate_all,
                      term_all) %>% 
apply(2,fivenum) -> summary_2

summary_2 <- t((summary_2))
colnames(summary_2) <- c("min", "p25","median","p75","max")

knitr::kable(summary_2,digits = 2, format.args = list(big.mark = ",",   scientific = FALSE))


```

Summary all time

```{r summary-all-time, comment="", message=FALSE, echo = FALSE, warning=FALSE}
  
data_r_all %>% 
  mutate(
  institution = ifelse(institution %in% banks, institution, "Other")) %>%  
  group_by(institution) %>% 
  summarise(N = n(),
            `Total Portfolio` = sum(total_portfolio )/1000000,
            `WA interest_rate` = sum(interest_rate_all * total_portfolio, na.rm = T) / sum(total_portfolio),
            `WA term` = sum(term_all * total_portfolio, na.rm = T) / sum(total_portfolio),
            `WA past_due_pct` = sum(past_due_pct*total_portfolio, na.rm = TRUE) / sum(total_portfolio)) %>% 
  mutate(`Total Portfolio (%)` = 100 * `Total Portfolio` / sum(`Total Portfolio`)) %>%  
  arrange(desc(`Total Portfolio (%)`)) -> table_1 

knitr::kable(table_1,digits = 2, format.args = list(big.mark = ",",   scientific = FALSE))


```

## Bivariate analysis

Loan amount by size of firm

```{r analysis-by-firm-size,  message=FALSE, echo = FALSE, warning=FALSE}

data_r_all %>% 
  mutate(firm_size_code = fct_relevel(firm_size_code,
                                      "S","L","T")) %>%
  ggplot(aes(x = log(total_portfolio), 
             color = firm_size_code) ) +
  geom_density() -> g1
data_r_all %>% 
  mutate(firm_size_code = fct_relevel(firm_size_code,
                                      "S","L","T")) %>% 
  ggplot(aes(x = log(total_portfolio),
                          y = firm_size_code)) + 
  geom_boxplot() + 
  coord_flip() -> g2
g1 + g2



```

\newpage

Loan amount by institution

```{r analysis-by-institution,  message=FALSE, echo = FALSE, warning=FALSE, fig.height = 8, fig.width = 13, fig.align = "center"}
data_r_all %>% 
  ggplot(aes(x = log(total_portfolio), 
             color = institution) ) +
  geom_density()  + 
  theme(legend.position = "bottom") 

data_r_all %>% 
  ggplot(aes(x = log(total_portfolio),
                          y = institution)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))




```

\newpage

Loan amount by SCIAN code

```{r analysis-by-sector,  message=FALSE, echo = FALSE, warning=FALSE, fig.height = 8, fig.width = 13, fig.align = "center"}
data_r_all %>% 
  ggplot(aes(x = log(total_portfolio), 
             color = scian_code) ) +
  geom_density()  + 
  theme(legend.position = "bottom")

data_r_all %>% 
  ggplot(aes(x = log(total_portfolio),
                          y = scian_code)) + 
  geom_boxplot() 

```

\newpage

Analysis by destination of funds

```{r analysis-by-fund-use,  message=FALSE, echo = FALSE, warning=FALSE, fig.height = 8, fig.width = 13, fig.align = "center"}
data_r_all %>% 
  ggplot(aes(x = log(total_portfolio), 
             color = factor(funds_use_code)) ) +
  geom_density()  + 
  theme(legend.position = "bottom")

data_r_all %>% 
  ggplot(aes(x = log(total_portfolio),
                          y = factor(funds_use_code)) ) + 
  geom_boxplot() 
```

## Share of banks across time

```{r share-of-banks, comment="", message=FALSE, echo = FALSE, warning=FALSE}

# function to calculate share for institution  
calc_share_f <- function(x){
  # x <- data_r_all
  x %>% group_by(institution) %>% 
    summarise("Number of loans" = sum(number_of_loans),
              "Total Portfolio" = sum(total_portfolio)) %>% 
    mutate(`Total Portfolio` = `Total Portfolio`/1000000000) %>% 
    mutate("Number of loans (%)" = 100*`Number of loans` / sum(`Number of loans`),
           "Total Portfolio (%)" = 100*`Total Portfolio` / sum(`Total Portfolio`)) %>% 
    arrange(desc(`Number of loans (%)`)) %>% 
    select(institution, `Number of loans`,
           `Total Portfolio`, `Number of loans (%)`, `Total Portfolio (%)`)
  }

# # check 
# data_r_all %>% filter(period == 201604) %>% 
#  calc_share_f()

# iterate over periods for G7 only 
data_r_all %>% mutate(
  institution = case_when(institution %in% banks ~ "G7",
                          !(institution %in% banks) ~ "other")) %>% 
  group_by(period) %>% nest() %>% 
  mutate(shares = map(.x = data, .f = calc_share_f)) %>% 
  unnest(shares) -> shares_g7_df
  
# iterate over periods for all banks 
data_r_all %>% mutate(
  institution = ifelse(institution %in% banks, institution, "Other")) %>% 
  group_by(period) %>% nest() %>% 
  mutate(shares = map(.x = data, .f = calc_share_f)) %>% 
  unnest(shares) -> shares_banks_df

# clean memory
#rm(data_r_all, data_r_all_nested)

# reshape for ggplot
reshape_shares_as_ts_f <- function(x){
  x %>% 
  select(period, institution,`Total Portfolio (%)`) %>% 
  pivot_wider(names_from = institution, 
              values_from = `Total Portfolio (%)`) %>% 
  select(-period) %>% 
  ts(start = c(2016,4),end = c(2021,11), frequency = 12) %>% 
  as_tsibble(index = "index", 
             pivot_longer = TRUE) %>% 
  filter(key != "period")
}

shares_g7_ts <- reshape_shares_as_ts_f(shares_g7_df)
shares_banks_ts <- reshape_shares_as_ts_f(shares_banks_df)

plot_shares_f <- function(data, faceting = FALSE){
  if(faceting == FALSE){
    data %>% 
      ggplot(aes(x = index, y = value, group = key, color = key)) +
      geom_line() + 
      ggtitle("G7 Banks shares", 
              subtitle = "Percent of Amount Outstanding") +
      ylab("Percentage (%)") + 
      xlab("Time")
  }else{
    data %>% 
      ggplot(aes(x = index, y = value, group = key, color = key)) +
      geom_line() + 
      facet_wrap( ~ key) + 
      ggtitle("G7 Banks shares", 
              subtitle = "Percent of Amount Outstanding") +
      ylab("Percentage (%)") + 
      xlab("Time")
  }
}

shares_g7_ts %>% plot_shares_f(faceting = FALSE)
shares_banks_ts %>% plot_shares_f(faceting = TRUE)


```
